---
title: "Introduction to tidyverse. Day 2."
output:
  html_document:
    toc: yes
    number_sections: yes
    theme: united
date: '2022-05-26'
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction into tidyverse

From the official website [tidyverse.org](https://www.tidyverse.org/): 

*"The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures."*

The most useful thing when learning the tidyverse is looking into their cheat sheets that include all important information about
the specific package. For example, [this is a cheat sheet for a dplyr package](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf).

*Remember, you always have to load library if you want to use some functions from there.*

## Install the complete tidyverse

```{r, eval = FALSE}
install.packages("tidyverse")
```

You can see what is installed using the command:
```{r, eval = FALSE}
library(tidyverse)
```

You can update all packages using the function `tidyverse_update()`.

## A tibble

Remember how we created the data frames using `data.frame()` function? In tidyverse, the data frames are called **tibbles**. 
They are much like the regular data frames, except **printing** and **subsetting**:

Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen. This makes it much easier to work with large data. In addition to its name, each column reports its type, a nice feature borrowed from `str()`:

```{r}
library(tibble)

tibble(
  id = seq(1:5),
  rock_name = c("charnockite", "pyroxenite", "dunite", "volynite", "anorthosite"),
  is.altered = c(TRUE, FALSE, FALSE, FALSE, FALSE)
)
```

You can convert a data frame into tibble by using function `as_tibble()`

```{r}
library(tibble)

df = data.frame(
  id = seq(1:5),
  rock_name = c("charnockite", "pyroxenite", "dunite", "volynite", "anorthosite"),
  is.altered = c(TRUE, FALSE, FALSE, FALSE, FALSE)
)

as_tibble(df)
```

# Importing the data using `readr` package

A tidyverse has a separate package for importing the data. Yesterday, we used base function to load the data frame - `read.table()`.
However, the tidyverse' package `readr` is much more flexible! Check [readr](https://readr.tidyverse.org/) documentation and a cheatsheet to know more.

```{r}
library(readr)
library(knitr)

initial_data <- readr::read_csv('./data/raw_data/mafic_dykes.csv', col_names=TRUE)
kable(initial_data[5:10,], caption='A sample table of XRF and ICP analyses of mafic dykes.')
```

Note, the data frame imported is already a tibble and all data types are "guessed" under the hood 🙂.

# Tidy data with `tidyr` 

The package called [tidyr](https://tidyr.tidyverse.org/) is used to create tidy data representation.

The problem is that the same data may be represented in **multiple ways** using the table representation. 
Here are some examples with same data but different representations:

```{r}
library(tidyr)

tibble(
  id = paste0(seq(1:5), '.sample'),
  mineral = c('magnetite', 'ilmenite', 'ilmenite', 'Cr-spinel', 'hercynite'),
  city = c('Banska Bystrica', 'Bratislava', 'Banska Bystrica', 'Banska Bystrica', 'Banska Bystrica'),
  institution = c('SAV', 'Geological Institute', 'SAV', 'SAV', 'SAV'),
  n_analyses = c(30, 119, 40, 20, 10)
)

tibble(
  id = paste0(seq(1:5), '.sample'),
  mineral = c('magnetite', 'ilmenite', 'ilmenite', 'Cr-spinel', 'hercynite'),
  lab = c('SAV, Banska Bystrica', 'Geological Institute, Bratislava', 'SAV, Banska Bystrica', 'SAV, Banska Bystrica', 'SAV, Banska Bystrica'),
  n_analyses = c(30, 119, 40, 20, 10)
)

tibble(
  city = c('Bratislava', 'Banska Bystrica'),
  institution = c('SAV', 'Geological Institute'),
  minerals = list(list('magnetite', 'ilmenite', 'ilmenite', 'Cr-spinel'), list('hercynite')),
  n_analyses = c(119, 100)
)
```

In order to make things clear, we'll use next terminology:

1. **A variable is a column.**
2. **An observation is a row.**
3. **A cell in the table is a value.**

**tidyr** functions fall into five main categories. We will not dive into those, because they are a bit complicated.

## Pivotting

Converts between long and wide forms. tidyr 1.0.0 introduces `pivot_longer()` and `pivot_wider()`. See `vignette("pivot")` for more details.

## Rectangling

Turns deeply nested lists (as from JSON) into tidy tibbles. See `unnest_longer()`, `unnest_wider()`, `hoist()`, and `vignette("rectangle")` for more details.

## Nesting

Converts grouped data to a form where each group becomes a single row containing a nested data frame, and unnesting does the opposite. See `nest()`, `unnest()`, and `vignette("nest")` for more details.

## Splitting and combining character columns

Use `separate()` and `extract()` to pull a single character column into multiple columns; use `unite()` to combine multiple columns into a single character column.

As an example of what these functions may do, let's look into the following:

```{r}
library(readr)
library(tidyr)
library(knitr)
library(dplyr)

initial_data <-readxl::read_xls('./data/raw_data/Chrysoberyl analysis database.xls')
kable(initial_data[1:5,], caption='A sample table of EPMA analyses of Be minerals.')
```

As you may see, this data is a bit **wide**.. We might want to convert it to a longer format, eg have the oxides as variables (columns)
and analyses as observations (rows) - that is more logical in our case. 
The latter could be done with regular function `t()` which transposes objects:
```{r}
initial_data_t <- t(initial_data)
colnames(initial_data_t) <- initial_data_t[1,]
initial_data_t <- as_tibble(cbind(Sample = names(initial_data)[-1], initial_data_t[-1,]))

initial_data_t <- 
  initial_data_t %>%
  mutate_at(vars(-Sample, -Origin, -Reference), as.numeric) %>%
  mutate_at(vars(-Sample, -Origin, -Reference), .funs=function(x){ replace(x, x<=0, NA) })

kable(initial_data_t[1:10,1:5], caption='A sample table of EPMA analyses of Be minerals.')
```

Now, looks a bit better and we are ready for some filtering and calculations!

# `dplyr` basics

[Dplyr](https://dplyr.tidyverse.org/) is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges.

Dplyr provides an alternative syntax for chaining or combining multiple functions, or calling a single function:

```{r}
select(initial_data_t, SiO2)
```
which is the same as this
```{r}
initial_data_t %>%
  select(SiO2)
```

**%>%** operator is called a **pipe** operator. Its syntax is very simple and it allows you to omit the `data` argument of tidyverse functions.

There are **5 most useful functions** within this package.

## Functions to work with observations (rows)

### `filter()`
It allows you to choose rows based on column values.

Let's filter our dataframe by those rows satisfying the criteria 
```{r}
initial_data_t %>%
  filter(Al2O3 > 50 & `BeO calc.` <= 19)
```
Note, the same could be written as:
```{r, eval = FALSE}
filter(initial_data_t, Al2O3 > 50 & `BeO calc.` <= 19)
```

To use filtering effectively, you have to know how to select the observations that you want using the comparison operators. R provides the standard suite: `>`, `>=`, `<`, `<=`, `!=` (not equal), and `==` (equal).

Multiple arguments to `filter()` are combined with “and”: every expression must be true in order for a row to be included in the output. For other types of combinations, you’ll need to use Boolean operators yourself: `&` is “and”, `|` is “or”, and `!` is “not”. Additionally,
there's a useful short-hand for selecting values satisfying multiple criteria using `x %in% y`. This will select every row where `x` is one of the values in `y`.

```{r}
initial_data_t %>% 
  filter(Origin %in% c('Magmatic leucogranite', 'Contaminated ultrabasic rock'))
```

#### Exercise

Try to do the following:

1. Open the file *exercises/exercise_2.R* and execute the code available there.
2. Now, try to filter the dataframe by those minerals, which have `BeO calc.` more or equal to 20wt% using the %>% operator and without it.
3. Add another filter to the previous one: filter out those rows which doesn't have a `Reference`. Hint: a function `is.na()` can help you check if the value is `NA`.

### `arrange()`
It allows you to change the order of the rows. This functions works just the same as `filter()` except that instead of selecting rows,
it changes their order. There's additional function `desc()` that could be used to arrange rows in descending order.

Let's arrange our dataframe by BeO content in ascending order:
```{r}
initial_data_t %>% 
  arrange(`BeO calc.`)
```
...and now in the descending:
```{r}
initial_data_t %>% 
  arrange(desc(`BeO calc.`))
```

We are also able to apply sorting using multiple columns:
```{r}
initial_data_t %>% 
  arrange(desc(`BeO calc.`), desc(TiO2), desc(Cr2O3))
```


### `slice()`
It allows you to choose rows based on location. It's kind of a helper function to choose some rows to work with.
```{r}
initial_data_t %>% 
  slice(5:10)
```

Additionally, you can use `slice_head(n=X)` and `slice_tail(n=X)` to select first X or last X rows respectively.

```{r}
initial_data_t %>% 
  slice_head(n=2)
```

`slice_sample(n=X)` will return a random X rows from a dataframe.
`slice_min()` and `slice_max()` select rows with highest or lowest values of a variable.

```{r}
initial_data_t %>% 
  slice_max(TiO2)
```

## Functions to work with variables (columns)

### `select()`
It allows you to choose rows based on column values.

More often than not, we want to work with only a small number of variables (columns). `select()` allows us to choose columns of 
interest!

We can choose one or multiple columns..
```{r}
initial_data_t %>% 
  select(Sample, TiO2, Cr2O3)
```

...a range of columns by using `:` operator:
```{r}
initial_data_t %>% 
  select(Sample, SiO2:Cr2O3)
```
...we can choose which columns NOT to include:
```{r}
initial_data_t %>% 
  select(Sample, !TiO2:Cr2O3)
```

But what is really helpful is that there's a bunch of helper functions to use with a `select()` function! Here are few of those:

- `starts_with()`
- `ends_with()`
- `matches()`
- `contains()`

the names of these functions already answer the question what each does, but here are few examples.
Let's select all columns that starts with the character 'Si' or character 'Ti':
```{r}
initial_data_t %>% 
  select(starts_with(c('Si', 'Ti')))
```
...now, let's select the cations from our dataset:
```{r}
initial_data_t %>% 
  select(!contains('O'))
```

### `rename()`
It allows you rename columns.

Often, we might need to rename our variables to a more appropriate name. Here we can use the `rename()` function!

```{r}
initial_data_t %>% 
  rename(silica = SiO2, alumina = Al2O3)
```

The latter returns the whole data frame with renamed columns.

### `mutate()`
It allows you to create columns.

This is a very handy method for calculating new columns.

```{r}
initial_data_t %>% 
  mutate(ti_to_be = TiO2/`BeO calc.`) %>%
  select(Sample, TiO2, `BeO calc.`, ti_to_be)
```
...we can calculate multiple columns at once:
```{r}
initial_data_t %>% 
  mutate(ti_to_be = TiO2/`BeO calc.`,
         cr_to_ti = Cr2O3/TiO2) %>%
  select(Sample, TiO2, Cr2O3, `BeO calc.`, ti_to_be, cr_to_ti)
```

## Functions to work with groups of observations (rows)

### `summarise()`
It allows you to collapse a group into a single row using a specific function as an argument.

For, example, we can calculate few stats for the whole data frame:
```{r}
initial_data_t %>% 
  summarise(silica_min = min(SiO2, na.rm=TRUE), min_alumina = min(Al2O3, na.rm=TRUE))
```

However, that's not very useful. Consider a task when we might want to calculate some stats based on the factors (or groups). For example,
in our case we want to see what is the mean value of FeO, Cr2O3, TiO2 and BeO for **each origin type**. There's where a `group_by()` comes in handy!

### `group_by()`

`group_by()` takes an existing tbl and converts it into a grouped tbl where operations are performed "by group". Let's see an example.

What happens if we do this?
```{r}
initial_data_t %>% 
  summarise(`BeO mean` = mean(`BeO calc.`))
```
...right. A mean value for all rows. But we want a mean BeO for each origin group! Let's do this:

```{r}
initial_data_t %>% 
  group_by(Origin) %>%
  summarise(`BeO mean` = mean(`BeO calc.`))
```
We may calculate as many variables per group as we wish:
```{r}
initial_data_t %>% 
  group_by(Origin) %>%
  summarise(`BeO mean` = mean(`BeO calc.`, na.rm=TRUE), 
            `FeO mean` = mean(FeO, na.rm=TRUE), 
            `Cr2O3 mean` = mean(Cr2O3, na.rm=TRUE), 
            `TiO2 mean` = mean(TiO2, na.rm=TRUE)
            )
```

As well as grouping by existing variables, like in the previous example, you can **group by any function of existing variables**. This is equivalent to performing a `mutate()` before the `group_by()`:
```{r}
initial_data_t %>% 
    mutate(Ti_group = if_else(TiO2 > 1, 'high', 'low')) %>%
    group_by(Ti_group) %>%
    summarise(`BeO mean` = mean(`BeO calc.`, na.rm=TRUE), 
              `FeO mean` = mean(FeO, na.rm=TRUE), 
              `Cr2O3 mean` = mean(Cr2O3, na.rm=TRUE), 
              `TiO2 mean` = mean(TiO2, na.rm=TRUE)
    )
```

# Exercises

Open a script `exercises/exercise_2.R`. You will use it to do the tasks!

## Task

Filter out the observations (rows), where Cr2O3 was not measured. Hint: you can use a helper function `is.na()` to test whether an object is NA. Assign this to a new variable called `filtered_`.

## Task

Filter the `filtered_` by those samples where TiO2 > 1wt%.

## Task

Now, arrange the `filtered_` data frame by `BeO calc.` in the descending order.

## Task

We need only few columns from `filtered_` to work with. Use a function `select()` to select the next variables:

- Sample
- Origin
- All columns that contain Oxygen ('O') in its formula.

## Task

Calculate the median of BeO, Al2O3, TiO2 and Cr2O3 by each Origin group of the `initial_data_t` data frame. Hint: use `group_by()` and `summarise()` functions.


# Data visualization with `ggplot`

Let's finally dive into visualizing the data!

We will be using [ggplot2](https://ggplot2.tidyverse.org/) package, which is a part of tidyverse ecosystem.

It’s hard to succinctly describe how **ggplot2** works because it embodies a deep philosophy of visualisation. However, in most cases you start with `ggplot()`, supply a dataset and aesthetic mapping (with `aes()`). You then add on layers (like `geom_point()` or `geom_histogram()`), scales (like `scale_colour_brewer()`), faceting specifications (like `facet_wrap()`) and coordinate systems (like `coord_flip()`).

See [the examples of ggplot2 visualizations](https://r-graph-gallery.com/) to know more.

Here's a simple example with the previous dataset.

```{r}
library(ggplot2)

ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = `BeO calc.`, y = TiO2))
```

The structure of code is as follows:

 - `ggplot()` - creates a new ggplot.
 - `aes()` - constructs [aesthetic mappings](https://ggplot2.tidyverse.org/reference/aes.html). Aesthetic mappings describe how variables in the data are mapped to visual properties (aesthetics) of geoms.
 - `+` - adds components to a plot.


Accordingly, a reusable template for making graphs with `ggplot2` would be:
```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
where 

- `DATA` is a tibble or a data frame, 
- `GEOM_FUNCTION` - how to visualize the data,
- `MAPPINGS` - which variables to use and how

What happens when we run `ggplot(data = initial_data_t)`? 

Open the [cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf) to start exploring this library.

## Scatter plot

### A basic one
Let's start with a simple scatter binary plot. Try to plot *TiO2* on X axis with *Al2O3* on Y axis for our dataframe!
```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = Al2O3))
```

### With custom color
Let's change the color of the points and adjust their size:

```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = Al2O3), color='blue', size=2)
```

### With custom stroke, shape and fill
Let's change the stroke of points, shape and a fill color:

```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = Al2O3), shape=21, color='blue', fill='white', size=2, stroke=1)
```

You can find the available options for R shapes [here](http://sape.inf.usi.ch/quick-reference/ggplot2/shape).

### With calculated colors
Let's assign colors depending on the Origin of our samples (`Origin` column). We can control this by passing `color` argument to `aes()` function.
```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = Al2O3, color=Origin), size=2)
```

#### Using `scale_color_brewer()`

Hmmm... you see that by default `ggplot2` assigns very similar colors to two groups - *Contaminated syenite pegmatite* and 
*Contaminated ultrabasic rock*. Luckily, `ggplot2` provides us with few functions to change the colors palette according to a
[discrete](https://ggplot2.tidyverse.org/reference/scale_brewer.html) or a [continuous](https://ggplot2.tidyverse.org/reference/scale_colour_continuous.html) scale.

We will append a function `scale_color_brewer()` to our `ggplot2` object and see what happens!
```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = Al2O3, color=Origin), size=2) +
  scale_color_brewer(palette = "Accent")
```

Try to pass `scale_color_brewer()` without any argument. What happens?
Now, try to use another palette. You can choose the palette of your preferences from [here](https://ggplot2.tidyverse.org/reference/scale_brewer.html#palettes).


### Create subplots, using `facet_wrap()`

One way to add additional variables is with aesthetics. Another way, particularly useful for **categorical variables**, is to split your plot into **facets** - *subplots that each display one subset of the data*.

To facet your plot by a single variable, use `facet_wrap()`. The first argument of `facet_wrap()` should be a formula, which you create with `~` followed by a variable name (here “formula” is the name of a data structure in R, not a synonym for “equation”). The variable that you pass to `facet_wrap()` should be discrete.

So, let's create a separate plot for each origin, using the facets!
```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = `BeO calc.`)) + 
  facet_wrap(~Origin)
```

You can even use a logical expression to create subplots. Consider an idea to create two subplots:

- BeO with TiO2 where BeO >= 18wt%
- BeO with TiO2 where BeO <= 18wt%

Here is how we can achieve that:
```{r}
ggplot(data = initial_data_t) +
  geom_point(mapping = aes(x = TiO2, y = `BeO calc.`)) + 
  facet_wrap(~ `BeO calc.` >= 18)
```

## Box plot

Often, we might want to see the spread of some data based on the group or a category. The useful visual representation of it can be a [box plot](https://r-graph-gallery.com/boxplot.html)!

### A basic plot
```{r}
ggplot(data = initial_data_t) +
  geom_boxplot(mapping = aes(x = Origin, y = `BeO calc.`, fill=Origin))
```

### Improve the labels
Since we are showing the legend, perhaps we might want to remove the X axis ticks labels, as they are too long and overlap.
We can control that by passing a function `scale_x_discrete()`.
Additionally, lets give some pretty labels to X and Y axis using the `xlab()` and `ylab()` functions.

```{r}
ggplot(data = initial_data_t) +
  geom_boxplot(mapping = aes(x = Origin, y = `BeO calc.`, fill=Origin)) + 
  scale_x_discrete(labels=NULL) +
  xlab(label="Host rock origins") +
  ylab(label='BeO (wt%)')
```

Looks better now!

### Create sub boxplots, using `facet_wrap()`

Just like we did previously with scatter plots, same logics is present when working with other plots.
Lets see what is the spread of TiO2 based on two groups:

- TiO2 where BeO >= 18wt%
- TiO2 where BeO <= 18wt%

However, now we will create a separate columns in our table, called `BeO_group` with two values in it:

- 'BeO >= 18 wt%', if BeO >= 18wt%
- 'BeO <= 18 wt%', if BeO <= 18wt%

We will use a `mutate()` method to do this.

```{r}
initial_data_t_ = initial_data_t %>%
  mutate(BeO_group = ifelse(`BeO calc.` >= 16, 'BeO >= 18 wt%', 'BeO <= 18 wt%'))

ggplot(data = initial_data_t_) +
  geom_boxplot(mapping = aes(x = Origin, y = TiO2, fill=Origin)) + 
  facet_wrap(~BeO_group) +
  scale_x_discrete(labels=NULL) +
  xlab(label="Host rock origins") +
  ylab(label='TiO2 (wt%)')
```

*Note: we have created a new data frame `initial_data_t_`. Typically, programmers append `_` to variable names to indicate
that these are temporary or internal ones.*


Now, lets create one plot, where *categories are organized in groups and subgroups*. Lets calculate another variable, called `Al2O3_group` with two values in it:

- 'Al2O3 >= 78 wt%', if Al2O3 >= 78 wt%
- 'Al2O3 <= 78 wt%', if Al2O3 <= 78 wt%

Now, lets plot `BeO` on Y axis, `Origin` on X axis and fill the box plots according to `Al2O3_group` variable.
```{r}
initial_data_t_ = initial_data_t %>%
  mutate(Al2O3_group = ifelse(Al2O3 >= 78, 'Al2O3 >= 78 wt%', 'Al2O3 <= 78 wt%'))

ggplot(data = initial_data_t_) +
  geom_boxplot(mapping = aes(x = Origin, y = `BeO calc.`, fill=Al2O3_group)) +
  scale_x_discrete(labels=NULL) +
  xlab(label="Host rock origins") +
  ylab(label='BeO (wt%)')
```

## Histogram

It is relatively straightforward to build a histogram with `ggplot2` thanks to the `geom_histogram()` function. Only one numeric variable is needed in the input. Note that a warning message is triggered with this code: we need to take care of the bin width as explained in the next section.

### A basic one

```{r}
ggplot(data = initial_data_t) +
  geom_histogram(mapping = aes(x = Al2O3))
```

### The `binwidth` parameter

A histogram takes as input a numeric variable and cuts it into several bins. Playing with the bin size is a very important step, since its value can have a big impact on the histogram appearance and thus on the message you’re trying to convey. 

```{r}
ggplot(data = initial_data_t) +
  geom_histogram(mapping = aes(x = Al2O3), binwidth = 0.5, fill='green', color='black')
```

What happens when you pass >1 as a `binwidth`?

### Multi histogram

Lets plot two histograms on same plot - BeO with Al2O3.

```{r}
ggplot(data = initial_data_t) +
  geom_histogram(mapping = aes(x = `BeO calc.`), binwidth = 0.5, fill='blueviolet', color='black') +
  geom_histogram(mapping = aes(x = Al2O3), binwidth = 0.5, fill='green', color='black')
```

As you see, the X axis label is a bit misleading.. Lets fix it using `xlab()` method.
```{r}
ggplot(data = initial_data_t) +
  geom_histogram(mapping = aes(x = `BeO calc.`), binwidth = 0.5, fill='blueviolet', color='black') +
  geom_histogram(mapping = aes(x = Al2O3), binwidth = 0.5, fill='green', color='black') +
  xlab(label='value (wt%)')
```

# Exercises

## Task

Build a binary plot with BeO on X axis and Cr2O3 on the Y axis. Do you see any patterns? Why? 

## Task

Now, change the color of each observation according to a sample country rock origins. In which rocks the trend is clearly visible? So why we observe that?

## Task

Change the color palette using the function `scale_color_brewer()`. Choose a palette you like from [here](https://ggplot2.tidyverse.org/reference/scale_brewer.html#palettes).

## Task

Now, change the label of X axis to `BeO (wt%)` and Y axis to `Cr2O3 (wt%)` using the functions `ylab()` and `xlab()`.

## Task

Build a density plot. Use `ggplot2` [documentation](https://ggplot2.tidyverse.org/reference/geom_density.html) for doing this. Additionally, you can use [these examples](https://r-graph-gallery.com/135-stacked-density-graph.html).

You will use a new function `geom_density()`. Use the Origin as a group for color, and any variable you wish for calculating densities.


# Congratulations 🙃

You passed the course! Cheers 🍺

